---
title: Metadata
description: Overview
author: 
  - name: Gavin Kelly
    affiliations: BABS - Francis Crick Institute
format:
  html:
    theme: cerulean
    fontsize: 1.1em
    linestretch: 1.7
code-fold: true
echo: false
code-summary: "Show the code"
categories: [Metadata]
params:
---
  

# Introduction

Load the data into a SummarizedExperiment which contains a matrix with
row and column annotation frames.


```{r}
#| label: intro
#| results: 'hide'
#| echo: false
#| message: false

library(tidyverse)
library(Hmisc)
library(emmeans)
library(ragg)
library(SummarizedExperiment)
library(ComplexHeatmap)
library(gt)
library(nlme)
library(openxlsx)
library(factoextra)
library(gridExtra)
library(compositions)
```

We're using the following settings:

```{r}
#| label: load
source("R/markdown.r", local = knitr::knit_global())
source("R/transforms.r", local = knitr::knit_global())
source("R/plot.r", local = knitr::knit_global())
source("R/output.r", local = knitr::knit_global())
source("R/input.r", local = knitr::knit_global())
fig_caption <- captioner()
set.seed(1)

knitr::opts_chunk$set(warning=FALSE, error=FALSE, message=FALSE,
                      dev=c("ragg_png","pdf"), out.width="100%",
                      fig.width=14, fig.height=10,
                      results='asis'
                      )
```


```{r}
#| label: load2
if (!isTRUE(getOption('knitr.in.progress'))) {
  params <- load_params("resources/Metadata", extra=TRUE)
} else {
  gt(data.frame(Setting=names(params),
                Value=unlist(params)))
}


dat <- read_my_files(params)
msg <- metadata(dat)$warnings

mcols(colData(dat)) <- data.frame(PCA=FALSE)
mcols(mcols(dat)) <- data.frame(PCA=FALSE)

metadata(colData(dat))$palette <- df2colorspace(colData(dat), "Set1")
metadata(rowData(dat))$palette <- df2colorspace(rowData(dat), "Set1")


```

`r msg`

```{r}
#| label: standardise

assay(dat, "standardised") <- t(scale(t(assay(dat, "raw")), center=TRUE, scale=TRUE))
metadata(dat)$assay$standardised <- "Standardised"

```

# Descriptive and visual summaries

Below, we summarise the observed values of our metadata. For
qualitative data, we show observed frequencies (in a table, when the
number of distinct values is moderate, otherwise in text form), along
with some stats on missingness.  For quantitative data, we show means,
some convenient quantiles (median is the 0.5 quantile). More
esoterically,  _Info_ is a measure of the likely impact of
observed ties/duplicates  - the lower the _Info_, the less efficient
any subsequent (non-parametric) tests are expected to be.  And _Gmd_
is a measure of variability that works well for far-from-normal data.

```{r}
#| label: metadata
#| fig-cap: ""
#| output: asis
Hmisc::html(Hmisc::describe(data.frame(colData(dat), check.names=FALSE)))

```
 
## Assay data

And now the same for the assay data:

```{r}
#| label: assaydata
#| fig-cap: ""
#| output: asis
Hmisc::html(Hmisc::describe(data.frame(t(assay(dat, "raw")), check.names=FALSE)))

```


# Terms Of Use

# Bibliography


