---
title: PCA
description: Principal components analysis
author: 
  - name: LSHTM-Crick team
format:
  html:
    theme: cerulean
    fontsize: 1.1em
    linestretch: 1.7
code-fold: true
echo: false
code-summary: "Show the code"
categories: [PCA]
params:
  row_distance: euclidean
  clustering_method_columns: ward.D2
  subject_vars: [SUBJECT_ACCESSION, STUDY_TIME_COLLECTED]
  value_vars: CONCENTRATION_VALUE_PREFERRED
  names_from: ANALYTE_PREFERRED
  input_file: data/datasets/SDY180/ResultFiles/mbaa_result.csv
---
  

# Introduction

Load the data into a SummarizedExperiment which contains a matrix with
row and column annotation frames.


```{r}
#| label: intro
#| results: 'hide'
#| echo: false
#| message: false

library(tidyverse)
library(Hmisc)
library(emmeans)
library(ragg)
library(SummarizedExperiment)
library(ComplexHeatmap)
library(gt)
library(nlme)
library(openxlsx)
library(factoextra)
library(gridExtra)
library(compositions)
library(here)
```

We're using the following settings:

```{r}
#| label: load
source("R/markdown.r", local = knitr::knit_global())
source("R/transforms.r", local = knitr::knit_global())
source("R/plot.r", local = knitr::knit_global())
source("R/output.r", local = knitr::knit_global())
source("R/input.r", local = knitr::knit_global())
fig_caption <- captioner()
set.seed(1)

knitr::opts_chunk$set(warning=FALSE, error=FALSE, message=FALSE,
                      dev=c("ragg_png","pdf"), out.width="100%",
                      fig.width=14, fig.height=10,
                      results='asis', fig.cap=expression(fig_caption())
                      )
```


```{r}
#| label: load2
if (!isTRUE(getOption('knitr.in.progress'))) {
  params <- load_params("resources/PCA", extra=TRUE)
} else {
  gt(data.frame(Setting=names(params),
                Value=unlist(lapply(params, toString))))
}

#dat <- read_my_files(params)
dat <- read_long_data(params, input_file=here::here(params$input_file))
dat <- read_extra(dat, params, here=here())
msg <- metadata(dat)$warnings


mcols(colData(dat)) <- data.frame(PCA=FALSE)
mcols(mcols(dat)) <- data.frame(PCA=FALSE)

metadata(colData(dat))$palette <- df2colorspace(colData(dat), "Set1")
metadata(rowData(dat))$palette <- df2colorspace(rowData(dat), "Set1")

col_legends <- lapply(
  names(metadata(colData(dat))$palette$Heatmap),
  function(md_name) {
    md <- metadata(colData(dat))$palette$Heatmap[[md_name]]
    if (is.function(md))
      Legend(col_fun=md, title=md_name)@grob
    else
      Legend(legend_gp=gpar(fill=md),labels=names(md), title=md_name)@grob
  }
)
sane_legends <- col_legends[]
tmp <- do.call(grid.arrange, c(sane_legends, ncol=ceiling(sqrt(length(sane_legends)))))
fig_caption("Sample metadata colour-codings")

row_legends <- lapply(
  names(metadata(rowData(dat))$palette$Heatmap),
  function(md_name) {
    md <- metadata(rowData(dat))$palette$Heatmap[[md_name]]
    if (is.function(md))
      Legend(row_fun=md, title=md_name)@grob
    else
      Legend(legend_gp=gpar(fill=md),labels=names(md), title=md_name)@grob
  }
)
sane_legends <- row_legends
tmp <- do.call(grid.arrange, c(sane_legends, ncol=ceiling(sqrt(length(sane_legends)))))
fig_caption("Feature metadata colour-codings")




```

`r msg`

```{r}
#| label: standardise

if ("impute" %in% names(params)) {
  assay(dat, "raw") <- impute_se(assay(dat,"raw"), params$"impute")
}

assay(dat, "standardised") <- t(scale(t(assay(dat, "raw")), center=TRUE, scale=TRUE))
metadata(dat)$assay$standardised <- "Standardised"

assay(dat, "log") <- log0_third(assay(dat, "raw"))
metadata(dat)$assay$standardised <- "Log"

assay(dat, "standardised-log") <- t(scale(t(assay(dat, "log")), center=TRUE, scale=TRUE))
metadata(dat)$assay$standardised <- "Standardised Log"

for (i in assayNames(dat)) {
  dat <- add_dim_reduct(dat, assay=i)
}


```

# Descriptive and visual summaries

PCA projects the data onto single-number summaries that start with the
most informative single number in PC1 and add the most
useful complementary dimensions consecutively.

```{r}
#| label: pca1
#| output: asis

ind <- which(names(colData(dat))==paste0(params$assay, ".PCA"))
pca <- data.frame(cbind(colData(dat)[!mcols(colData(dat))$PCA], colData(dat)[, ind]), check.names=FALSE)

ojs_define(pca=pca)
ojs_define(colourBys=names(colData(dat))[!mcols(colData(dat))$PCA])
ojs_define(PCBys=names(colData(dat)[[ind]]))

mat <- data.frame(cbind(colData(dat)[!mcols(colData(dat))$PCA], data.frame(t(assay(dat, params$assay)), check.names=FALSE)), check.names=FALSE)
ojs_define(mat=mat)
ojs_define(features=row.names(dat))

cols <- metadata(colData(dat))$palette$ggplot
is_factor <- sapply(cols, function(x) !is.null(names(x)))
js_cols <- mapply(
  function(rgb, is_fac, cdat) {list(
    domain=if (is_fac) names(rgb) else range(cdat, na.rm=TRUE),
    range=rgb,
    type=ifelse(is_fac,"categorical", "linear"))},
  cols,
  is_factor,
  data.frame(colData(dat)[names(cols)]),
  SIMPLIFY=FALSE)
ojs_define(colscale_list=js_cols)

```

```{ojs}
viewof mycolor = Inputs.select(colourBys, {value: "region", label: "Colour By:"})
viewof myx = Inputs.select(PCBys, {value: "PC1", label: "Horizontal PC:"})
viewof myy = Inputs.select(PCBys, {value: "PC2", label: "Vertical PC:"})

Plot.dot(
    transpose(pca),
    {x: d => d[myx], y: d => d[myy], stroke: d => d[mycolor], fill: d => d[mycolor]}
  ).plot({
    color: {type: colscale_list[mycolor].type,
          domain: colscale_list[mycolor].domain,
		  range:  colscale_list[mycolor].range
		  },
    x: {label: myx},
	y: {label: myy}
  }
)

```

```{ojs}
viewof myxfeat = Inputs.select(features, {label: "Horizontal Feature:"})
viewof myyfeat = Inputs.select(features, {label: "Vertical Feature:"})
Plot.dot(
  transpose(mat),
      {x: d => d[myxfeat], y: d => d[myyfeat], stroke: d => d[mycolor], fill: d => d[mycolor]}
	  ).plot({
    color: {type: colscale_list[mycolor].type,
          domain: colscale_list[mycolor].domain,
		  range:  colscale_list[mycolor].range
		  },
    x: {label: myx},
	y: {label: myy}
  })
```

## Principal Cytokine:Covariate association


```{r}
#| label: pcacovar
#| output: asis

pc <- data.frame(colData(dat)[, names(colData(dat))==paste0(params$assay, ".PCA")], check.names=FALSE)
fitFrame <- colData(dat)
yvar <- make.unique(c(colnames(fitFrame), "y", sep = ""))[ncol(fitFrame) + 1]
fml <- update( ~ GENDER + RACE + MAX_SUBJECT_AGE + STUDY_TIME_COLLECTED + ARM_ACCESSION,
              paste(yvar, "~ ."))
plotFrame <- expand.grid(
  Covariate = attr(terms(fml), "term.labels"),
  PC = 1:ncol(pc))
plotFrame$Assoc <- NA
npc <- ncol(pc)
fit_selected <- list()
for (ipc in 1:npc) {
  fitFrame[[yvar]] <- pc[, ipc]
  fit1 <- lm(fml, data = fitFrame)
  fit_selected[[ipc]] <- MASS::stepAIC(fit1, trace = 0)
  ind <- attr(terms(fit_selected[[ipc]]), "term.labels")
  if (length(ind)) {
    fit0 <- as.data.frame(anova(fit_selected[[ipc]]))
    rss <- fit0[ind, "Sum Sq"]/sum(fit0[, "Sum Sq"])
    ind_pc <- plotFrame$PC == ipc
    ind_fit <- match(ind, plotFrame$Covariate[ind_pc])
    plotFrame$Assoc[ind_pc][ind_fit] <- rss
  }
}
plotFrame$wrap <- (plotFrame$PC - 1)%/%20
plotFrame$wrap <- paste0("PCs ", plotFrame$wrap * 20 +
                                   1, "-", min((plotFrame$wrap + 1) * 20, npc))
plotFrame$PC <- sprintf("%02d", plotFrame$PC)
pl <- ggplot(plotFrame, aes(x = PC, y = Covariate, fill = Assoc)) +
  geom_raster() +
  facet_wrap(~wrap, scales = "free_x",ncol = 1) +
  scale_fill_gradient2(low = "#4575b4",mid = "grey90", high = "#d73027") +
  theme_bw() +
  theme(aspect.ratio = length(unique(plotFrame$Covariate))/min(20, npc))
print(pl)
fig_caption("Metadata association with PCs")
  
```

# Terms Of Use


# Bibliography


